---
title: "Dorm to dorm"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Correlations between dorms and methods

Libraries.
```{r libs}
R.Version()$version.string
tellme <- function(name){message("Package ", name, " version: ", packageVersion(name))}
library(lubridate); tellme("lubridate")
library(dplyr); tellme("dplyr")
library(tidyr); tellme("tidyr")
library(ggplot2); tellme("ggplot2")
library(Hmisc); tellme("Hmisc")
library(ggrepel); tellme("ggrepel")
```

Functions
```{r}
# much of this is taken from
# http://www.sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization

#functions
tidyMelt <- function(cormat, values_name="values"){
    cormat.new = cormat %>% data.frame()
    cormat.new$var1 = row.names(cormat.new)
    melted_cormat = cormat.new %>% pivot_longer(cols=-var1, names_to = "var2", values_to = values_name)
    return(melted_cormat)
}

make_heatmap_1 <- function(cor.co2){
    ggplot(data = cor.co2, aes(x=var1, y=var2, fill=cor)) + 
        geom_tile(color = "white")+
        scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                             midpoint = 0, limit = c(-1,1), space = "Lab", 
                             name="Correlation",
                             na.value = gray(.9)) +
        theme_minimal()+ # minimal theme
        xlab("") +
        ylab("") +
        theme(axis.text.x = element_text(angle = 90, vjust = .5,
                                         size = 8, hjust = 1)) +
        coord_fixed()
}

# put a correlation matrix in long form without duplicate values
gatherCorMatrix <- function(corMat){
    corUpperTri = corMat
    corUpperTri[!upper.tri(corUpperTri)] = NA
    corLong = corUpperTri %>%
        as.data.frame() %>%
        mutate(rowName=row.names(corUpperTri)) %>%
        gather(key=colName, value=cor, -rowName) %>%
        filter(!is.na(cor))
    return(corLong)
}

# this trick stolen from: 'Tom Constant's answer here: https://stackoverflow.com/questions/10762287/how-can-i-format-axis-labels-with-exponents-with-ggplot2-and-scales
scientific_10 <- function(x) {
  parse(text=gsub("e[+]*", " %*% 10^", scales::scientific_format()(x)))
}
```

Theme.
Set color constants for the year.
```{r}
# based on ggplot defaults
yearColors = c('2021'="#00BA38", '2022'="#F8766D", '2023'="#619CFF")
yearPch = c('2021'=15, '2022'=16, '2023'=17)
```

Output
```{r}
outDir="../output"
if (!dir.exists(outDir)) dir.create(outDir)
```

## Input data

```{r}
rna.wide = read.delim("../../input/from_ncat/SARS-CoV-2_RNA-copies/RNA_copies_2021-2023.txt") %>%
    mutate(date = mdy(date, tz="")) 
rna = rna.wide %>% 
    gather(key=dorm, value=RNA.copies, Aggie.V4:Aggie.V6) %>%
    filter(!is.na(RNA.copies))
rna.wide = spread(rna, key=dorm, value=RNA.copies) #new values are numeric
head(rna)
```

### Correltation between sampling methods

Note on methods:
AggieV4 hall and Barbee.
grab sampling -- all non-AS samples
compound sampling --> AS auto-sampling

```{r}
rna.wide.stringy = rna.wide %>% 
    mutate(day = day(date)) %>%
    mutate(month = month(date, label=T)) %>%
    mutate(year = as.character(year(date))) %>%
    mutate(dateString=paste(day, month, year, sep="-"))
```

Plot Aggie V4 values.
```{r}
aV4test = cor.test(x=rna.wide$Aggie.V4, y=rna.wide$Aggie.V4.AS, use="complete", method="spearman")
aV4testN = sum(!is.na(rna.wide$Aggie.V4) & !is.na(rna.wide$Aggie.V4.AS))

ggplot(rna.wide.stringy, 
       aes(x=Aggie.V4, y=Aggie.V4.AS, label=dateString, col=year, pch=year)) +
    scale_y_log10(label=scientific_10) +
    scale_x_log10(label=scientific_10) +
    scale_color_manual(values=yearColors) +
    scale_shape_manual(values=yearPch) +
    xlab("Grab Sampling") +
    ylab("Auto Sampling") +
    ggtitle("Aggie V4 - comparative sampling methods", sub="RNA copies per liter") +
    geom_abline(slope=1, intercept = 0, alpha=.2) +
    geom_point() +
    geom_text_repel(show.legend = F) +
    labs(caption = paste0("Spearmans rho: ", round(aV4test$estimate, 3), 
                          "\n", "p-value: ", aV4test$p.value, "; \u2212log10(p): ", -log10(aV4test$p.value),
                          "\n", "Gray line is unity (x=y).", 
                          "\n", "N=", aV4testN)) +
    theme_classic()
ggsave("../output/sampleMethod_AggieV4.png", width = 7, height = 7)
```

Plot Barbee values.
```{r}
barbeeTest = cor.test(x=rna.wide$Barbee, y=rna.wide$Barbee.AS, use="complete", method="spearman")
barbeeTestN = sum(!is.na(rna.wide$Barbee) & !is.na(rna.wide$Barbee.AS))

ggplot(rna.wide.stringy, 
       aes(x=Barbee, y=Barbee.AS, label=dateString, col=year, pch=year)) +
    scale_y_log10(label=scientific_10) +
    scale_x_log10(label=scientific_10) +
    scale_color_manual(values=yearColors) +
    scale_shape_manual(values=yearPch) +
    xlab("Grab Sampling") +
    ylab("Auto Sampling") +
    ggtitle("Barbee - comparative sampling methods", sub="RNA copies per liter") +
    geom_abline(slope=1, intercept = 0, alpha=.2) +
    geom_point() +
    geom_text_repel(show.legend = F) +
    labs(caption = paste0("Spearmans rho: ", round(barbeeTest$estimate, 3), 
                          "\n", "p-value: ", barbeeTest$p.value, "; \u2212log10(p): ", -log10(barbeeTest$p.value),
                          "\n", "Gray line is unity (x=y).",
                          "\n", "N=", barbeeTestN)) +
    theme_classic()

ggsave("../output/sampleMethod_Barbee.png", width = 7, height = 7)
```


These two sampling approaches produce very significantly correlated results.  

It is still possible that one or the other method is more effective at predicting clinical outcomes (perhaps even significantly).  



### correlation between dorms

How correlated are the dorms?

```{r}
table(rna$dorm)
```

```{r}
corList.spearman = rcorr(rna.wide %>% select(-date) %>% as.matrix(),type="spearman")

write.my.table <- function(df, file){
    write.table(x=cbind(drom=row.names(df), df),
                file=file,
                sep="\t", quote=F, row.names=F)
}

corList.spearman$r %>% tidyMelt(values="cor") %>% make_heatmap_1() +
    scale_fill_gradient2(low = "red", high = "blue", mid = "white",
                             midpoint = .8, limit = c(0,0.9999), space = "Lab",
                             na.value = "gray30") +
    ggtitle("Spearman Correlation Coefficient")
ggsave("../output/dorm-ww-cor-matrix_rho.png")
write.my.table(corList.spearman$r, "../output/dorm-ww-cor-matrix_rho.txt")

negLogPvals = -log10(corList.spearman$P)
negLogPvals %>% tidyMelt(values="cor") %>% make_heatmap_1() +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                             midpoint = -log10(0.05), limit = c(0,max(negLogPvals)), space = "Lab",
                             name="\u2212log10(p)",
                             na.value = "gray30") +
    ggtitle("Spearman - p-values")
ggsave("../output/dorm-ww-cor-matrix_p.png")
write.my.table(corList.spearman$r, "../output/dorm-ww-cor-matrix_p.txt")

corList.spearman$n %>% tidyMelt(values="cor") %>% make_heatmap_1() +
    scale_fill_gradient2(low = "blue", high = "orange", mid = "white",
                             midpoint = 0, limit = c(0,max(corList.spearman$n)), space = "Lab",
                             name="N",
                             na.value = "gray30") +
    ggtitle("Number pairwise complete observations")
ggsave("../output/dorm-ww-cor-matrix_n.png")
write.my.table(corList.spearman$r, "../output/dorm-ww-cor-matrix_n.txt")
```


```{r}
# corUpperTri = corList.spearman$r
# corUpperTri[!upper.tri(corUpperTri)] = NA
# corLong = corUpperTri %>%
#     as.data.frame() %>%
#     mutate(rowName=row.names(corUpperTri)) %>%
#     gather(key=dorm, value=cor, -rowName) %>%
#     filter(!is.na(cor))

corLong = gatherCorMatrix(corList.spearman$r)


summary(corLong$cor)
```

We have one case with a very low correlation.  This Aggie.V6 and Holland, which is the intersection of the two dorms with the fewest samples.

```{r}
corLong %>% filter(cor == min(cor))
```

Excluding that pair of dorms, all dorms had pretty good correlation values.
```{r}
nextMin = corLong %>% filter(cor != min(cor)) %>% select(cor) %>% min()
nextMin
nextNextMin = corLong %>% filter(cor != min(cor)) %>% filter(cor != min(cor)) %>% select(cor) %>% min()
nextNextMin
```

The next two lowest values are around .5.

```{r}
hist(corLong$cor, breaks=40, col="lightblue", las=1, xlim=c(0,1), main="between dorm correlations", xlab="Spearmans Rho")
```

```{r}
# pdf("../output/between-dorm-correlations.pdf", width = 5, height = 4)
png("../output/between-dorm-correlations.png", width = 5, height = 4, units = "in", res=200)
hist(corLong$cor, breaks=40, col="lightblue", las=1, xlim=c(0,1), main="between dorm correlations", xlab="Spearmans Rho")

dev.off()

png("../output/between-dorm-correlations_with-annotation.png", width = 5, height = 4, units = "in", res=200)

hist(corLong$cor, breaks=40, col="lightblue", las=1, xlim=c(0,1), main="between dorm correlations", xlab="Spearmans Rho")
abline(v=barbeeTest$estimate, lwd=3)
abline(v=aV4test$estimate, lwd=3)
mtext(side=4, line=1,
      paste0("reference lines - correlation of AS-Grab. \nBarbee: ", round(barbeeTest$estimate,3), "; Aggie V4: ", round(aV4test$estimate,3)))
dev.off()
```

The correlation values for the between-method correlations for the same building are shown for reference.

Barbee: ``r barbeeTest$estimate``

Aggie V4: ``r aV4test$estimate``

```{r}
corLong %>% arrange(cor) %>% head(n=10)
# corLong %>% filter(rowName=="Aggie.V1" | rowName=="Aggie.V3") %>% filter(dorm=="Aggie.V1" | dorm=="Aggie.V3")
```

Haley Hall and Pride Hall are very close together on the NCA&T map. https://www.ussportscamps.com/pdf/running/ncat_campusmap.pdf

So are Barbee and Morrow, but on the other end of compus.




