---
title: "Campus-wide-cases"
output: html_document
date: "2024-09-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

Libraries.
```{r}
R.Version()$version.string
tellme <- function(name){message("Package ", name, " version: ", packageVersion(name))}
library(lubridate); tellme("lubridate")
library(dplyr); tellme("dplyr")
library(tidyr); tellme("tidyr")
library(ggplot2); tellme("ggplot2")
library(Hmisc); tellme("Hmisc")
library(ggrepel); tellme("ggrepel")

library(scales)
```


### color constants


Some data is given by dorm.  Set some color constants for the dorm buildings.
```{r}
dormColors = c(Aggie.V4="#6A3D9A",
               Aggie.V4.AS="#CAB2D6",
               Aggie.V1="gold4",
               Aggie.V3="gray20",
               Aggie.VE="#B15928",
               Aggie.V6="#FF7F00",
               Barbee="#33A02C",
               Barbee.AS="#B2DF8A",
               Morrow="#FB9A99",
               Holland="#FDBF6F",
               Haley="#E31A1C",
               Cooper="#1F78B4",
               Pride="#A6CEE3")
```

Set color constants for spring and fall terms.
```{r}
termSeasonPalette=c(fall="gold", spring="darkolivegreen3")
```

Set color constants for the year.
```{r}
# based on ggplot defaults
yearColors = c('2021'="#F8766D", '2022'="#00BA38", '2023'="#619CFF")
```


Set other color constants.
```{r}
sarsColor = "coral"
pmmColor = "dodgerblue"
ratioColor = "pink"
rnaTypeColorPalette = c(sarsColor = "coral", pmmColor="dodgerblue", ratioColor="pink")
```

Output
```{r}
outDir="../output"
if (!dir.exists(outDir)) dir.create(outDir)
```


## Clinical Cases

Read the number of cases per month.  Parse dates as date objects.  The parsed dates will be given as the day at the end of the month, because I assume that is the reporting time for the month as a whole. However, the plotting package (ggplot2) seems to want to draw the January line at January first.

```{r}
cc = read.delim("../../input/from_ncat/campus_clinical-cases/clinical_cases_2021-2023.txt") %>% 
    select(month, clinical.cases) %>% 
    filter(!is.na(month)) %>% 
    filter(!is.na(clinical.cases)) 
cc$time.month.start = floor_date(dmy(paste0("15-", cc$month), tz=""), unit = "month") %>% as.POSIXct()
cc$time.month.end = ceiling_date(dmy(paste0("15-", cc$month), tz=""), unit = "month") %>% as.POSIXct()
head(cc,3)
```

This is a table of ``r nrow(cc)`` rows, with dates ranging from ``r min(cc$time.month)`` to ``r max(cc$time.month)``.









## RNA copies


```{r}
rna.wide = read.delim("../../input/from_ncat//SARS-CoV-2_RNA-copies/RNA_copies_2021-2023.txt") %>%
    mutate(date = mdy(date, tz="")) 
rna = rna.wide %>% 
    gather(key=dorm, value=RNA.copies, Aggie.V4:Aggie.V6) %>%
    filter(!is.na(RNA.copies))
rna.wide = spread(rna, key=dorm, value=RNA.copies) #new values are numeric
head(rna)
```

This is a table of ``r nrow(rna.wide)`` rows, with dates ranging from ``r min(rna$date)`` to ``r max(rna$date)``.




## PMMoV by dorm

I wonder if our total lack of corralation between PMMoV and SARS may actually be because we have collapsed the data by across unrelated samples.  So I asked Shilpi for the per-drom data.

```{r}
pmmD.2022 = read.delim("../../input/from_ncat/PMMoV/PMMoV_2022_by-dorm_2024-07-08.txt")
pmmD.2023 = read.delim("../../input/from_ncat/PMMoV/PMMoV_2023_by-dorm_2024-07-08.txt")

# only 2022 has data for Aggie.V6, and only a few dates; drop the Aggie.V6 columns.
iNames = intersect(names(pmmD.2022), names(pmmD.2023))

pmmD = rbind(pmmD.2022[,iNames], pmmD.2023[,iNames]) %>%
    gather(key="dorm.rep", value="PMMoV", -date) %>%
    mutate(date = mdy(date, tz="")) %>%
    mutate(dorm=sub(x=dorm.rep, pattern=".rep[123]",replacement="")) %>%
    mutate(replicate=gsub(x=dorm.rep, pattern="[^?]*rep", replacement="rep")) %>%
    select(date, PMMoV, dorm, replicate) %>%
    spread(key=replicate, value=PMMoV) 

head(pmmD, 3)
```

```{r}
pmmD$mean = apply(pmmD[,c("rep1", "rep2", "rep3")], MARGIN = 1, FUN=mean)
pmmD$median = apply(pmmD[,c("rep1", "rep2", "rep3")], MARGIN = 1, FUN=median)
pmmD$sd = apply(pmmD[,c("rep1", "rep2", "rep3")], MARGIN = 1, FUN=sd)
head(pmmD, 3)
```

A little sanity check... does rep1 predict rep2 (and other pairings?).  It looks like once in a while one rep is very different.  Insead of averaging this in, we could take the median value across the three replicates.  This probably gives us a pretty similar value...right?

```{r}
ggplot(pmmD, aes(x=rep1, y=rep2)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()
ggplot(pmmD, aes(x=rep3, y=rep2)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()
ggplot(pmmD, aes(x=rep1, y=rep3)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()

ggplot(pmmD, aes(x=mean, y=median)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()
```

As expected, replicates have very similar values and taking either the median or the mean should give us very similar results.


Compare this to the per-dorm RNA copies for SARS-CoV-2.

```{r}
pmmD = pmmD %>% mutate(PMMoV = median) 
```












### PMMoV vs SARS per dorm 

merge and normalize by pmm

```{r}
pmm.merged = merge(x=rna %>% rename(SARS=RNA.copies),
                   y=pmmD %>% select(date, dorm, PMMoV)) %>%
    mutate(SARSperPMM.ratio = SARS / PMMoV) %>%
    mutate(SARSperPMM.logRatio = log10(SARS) / log10(PMMoV)) %>%
    mutate(month = round_date(date, "month"))
head(pmm.merged,3)
```

```{r}
pmm.merged.summary = pmm.merged %>%
    group_by(month) %>%
    summarise(mean.logRatio=mean(SARSperPMM.logRatio, na.rm=T), 
              mean.ratio=mean(SARSperPMM.ratio, na.rm=T), 
              # max.RNA.copies=max(RNA.copies), 
              # median.RNA.copies=median(RNA.copies), 
              # min.RNA.copies=min(RNA.copies), 
              n.vals=n()) %>%
    mutate(year=as.factor(lubridate::year(month)))
head(pmm.merged.summary)
```



### Campus case counts vs RNA.copies

Get RNA copies per month across the 3-year period
```{r}
rna.wide.month = rna.wide %>% mutate(month=round_date(date, unit="month"))

rna.long.month = rna.wide.month %>% 
    select(-date) %>% 
    gather(key=dorm, value=RNA.copies, -month) %>%
    filter(!is.na(RNA.copies)) %>%
    group_by(month) %>%
    summarise(mean.RNA.copies=mean(RNA.copies), 
              meanLog.RNA.copies=mean(log10(RNA.copies)), 
              max.RNA.copies=max(RNA.copies), 
              median.RNA.copies=median(RNA.copies), 
              min.RNA.copies=min(RNA.copies), 
              n.RNA.copies.vals=n()) %>%
    mutate(year=as.factor(lubridate::year(month)))

ggplot(rna.long.month, aes(x=mean.RNA.copies, y=max.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")
ggplot(rna.long.month, aes(x=mean.RNA.copies, y=median.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")
ggplot(rna.long.month, aes(x=min.RNA.copies, y=median.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")

ggplot(rna.long.month, aes(x=mean.RNA.copies, y=meanLog.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")
ggplot(rna.long.month, aes(x=max.RNA.copies, y=meanLog.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")

```

Merge in campus case counts by month, across the three year period.
```{r}
cc.month = cc %>% select(-month) %>% rename(month = time.month.start)
campus.cc.rna = merge(cc.month, 
                      rna.long.month, by="month")
head(campus.cc.rna)
```

```{r}
rm(ct)
ct=cor.test(x=campus.cc.rna$mean.RNA.copies, y=campus.cc.rna$clinical.cases, use="complete", method="spearman")
ggplot(data=campus.cc.rna, aes(x=mean.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()

rm(ct)
ct=cor.test(x=campus.cc.rna$median.RNA.copies, y=campus.cc.rna$clinical.cases, use="complete", method="spearman")
campus.wide.rna.mean.plot = ggplot(data=campus.cc.rna, aes(x=median.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()
campus.wide.rna.mean.plot

rm(ct)
ct=cor.test(x=campus.cc.rna$max.RNA.copies, y=campus.cc.rna$clinical.cases, use="complete", method="spearman")
ggplot(data=campus.cc.rna, aes(x=max.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()

rm(ct)
ct=cor.test(x=campus.cc.rna$meanLog.RNA.copies, y=campus.cc.rna$clinical.cases, use="complete", method="spearman")
ggplot(data=campus.cc.rna, aes(x=meanLog.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()
```

Using the mean values, put data on a log scale.
```{r}
ggplot(data=campus.cc.rna, aes(x=median.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values", subtitle = "x on log10") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()

ggplot(data=campus.cc.rna, aes(x=median.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    ggtitle("NCA&T campus wide values", subtitle = "no log scale") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()

ggplot(data=campus.cc.rna, aes(x=median.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("NCA&T campus wide values", subtitle = "x and y log10") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()
```
Only include the 22 and 23 data
```{r}
campus.cc.rna.2223 = campus.cc.rna %>% filter(year == "2022" | year == "2023")
ct2223=cor.test(x=campus.cc.rna.2223$median.RNA.copies, y=campus.cc.rna.2223$clinical.cases, use="complete", method="spearman")


ggplot(data=campus.cc.rna.2223, 
       aes(x=median.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values", subtitle = "x on log10") +
    labs(caption=paste0("Spearmans rho: ", round(ct2223$estimate, 3),
                        "\nCor test p-value: ", round(ct2223$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()

ggplot(data=campus.cc.rna.2223, 
       aes(x=median.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("NCA&T campus wide values", subtitle = "x and y log10") +
    labs(caption=paste0("Spearmans rho: ", round(ct2223$estimate, 3),
                        "\nCor test p-value: ", round(ct2223$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()
```
# PMMoV

```{r}
# pmm = read.delim("../../input/from_ncat/PMMoV/PMMoV_SARS_2022-2023.txt") %>%
#     select(date, SARS.CoV.2.means, PMMoV.means, ratio.SARStoPMMoV) %>%
#     mutate(date = mdy(date, tz="")) %>%
#     mutate(logRatioS2P = log10(SARS.CoV.2.means) / log10(PMMoV.means))
# head(pmm, 3)
```


## by pmmov

```{r}
head(pmm.merged.summary)

campus.cc.pmm = merge(cc.month, 
                      pmm.merged.summary, by="month")
head(campus.cc.pmm)
```

```{r}
rm(ct)
ct=cor.test(x=campus.cc.pmm$mean.logRatio, y=campus.cc.pmm$clinical.cases, use="complete", method="spearman")

ggplot(data=campus.cc.pmm, aes(x=mean.logRatio, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()
ggsave("../output/SARS-to-PMMoV-ratio_vs_campus-clinical-cases.png")

last_plot() + scale_y_log10()



rm(ct)
ct=cor.test(x=campus.cc.pmm$mean.ratio, y=campus.cc.pmm$clinical.cases, use="complete", method="spearman")

ggplot(data=campus.cc.pmm, aes(x=mean.ratio, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()

last_plot() + scale_y_log10()
```




```{r}
sessionInfo()
```




