---
title: "Cinical-cases"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Setup

Libraries.
```{r}
R.Version()$version.string
tellme <- function(name){message("Package ", name, " version: ", packageVersion(name))}
library(lubridate); tellme("lubridate")
library(dplyr); tellme("dplyr")
library(tidyr); tellme("tidyr")
library(ggplot2); tellme("ggplot2")
# library(Hmisc); tellme("Hmisc")
# library(ggrepel); tellme("ggrepel")
# 
library(scales)
```

Set color constants for spring and fall terms.
```{r}
termSeasonPalette=c(fall="gold", spring="darkolivegreen3")
```

Some data is given by dorm.  Set some color constants for the dorm buildings.
```{r}
dormColors = c(Aggie.V4="#6A3D9A",
               Aggie.V4.AS="#CAB2D6",
               Aggie.V1="gold4",
               Aggie.V3="gray20",
               Aggie.VE="#B15928",
               Aggie.V6="#FF7F00",
               Barbee="#33A02C",
               Barbee.AS="#B2DF8A",
               Morrow="#FB9A99",
               Holland="#FDBF6F",
               Haley="#E31A1C",
               Cooper="#1F78B4",
               Pride="#A6CEE3")
dormPch = 1:length(dormColors) + 10
names(dormPch) = names(dormColors)
```

Set color constants for the year.
```{r}
# based on ggplot defaults
yearColors = c('2021'="#00BA38", '2022'="#F8766D", '2023'="#619CFF")
yearPch = c('2021'=15, '2022'=16, '2023'=17)
```


# NCA&T Reference

I got the dates for the first and last day of classes for spring and fall sessions for the terms corresponding to the NCA&T data.

```{r}
classDates = read.delim("../../input/from_www/ncat_class-sessions.txt", comment.char = "#") %>%
    select(term, season, classesStart, classesEnd) %>%
    # when parsing the date, use tz="" so the resulting date object is a POSIXct object to be compatible with ggplot2
    mutate(classesStart = dmy(classesStart, tz=""))  %>%
    mutate(classesEnd = dmy(classesEnd, tz="")) 
classDates
```

# Data from NCS&T

This is data that was emailed to Ivory Blakley by Shilpi Bhatia.

## Clinical Cases

Read the number of cases per month.  Parse dates as date objects.  The parsed dates will be given as the day at the end of the month, because I assume that is the reporting time for the month as a whole. However, the plotting package (ggplot2) seems to want to draw the January line at January first.

```{r}
cc = read.delim("../../input/from_ncat/campus_clinical-cases/clinical_cases_2021-2023.txt") %>% 
    select(month, clinical.cases) %>% 
    filter(!is.na(month)) %>% 
    filter(!is.na(clinical.cases)) 
cc$time.month.start = floor_date(dmy(paste0("15-", cc$month), tz=""), unit = "month") %>% as.POSIXct()
cc$time.month.end = ceiling_date(dmy(paste0("15-", cc$month), tz=""), unit = "month") %>% as.POSIXct()
head(cc,3)
```

This is a table of ``r nrow(cc)`` rows, with dates ranging from ``r min(cc$time.month)`` to ``r max(cc$time.month)``.

Plot cases over time. show each month as a time range rather than a time point so as to clearly show what time range that data represents. Use shaded rectangles to indicate the time range of classes for spring and fall semesters.

```{r}
ggplot(cc, aes(x=time.month.start, y=clinical.cases)) +
    xlab("time (month)") +
    scale_x_datetime(labels = date_format("%b%y")) +
    # scale_x_continuous(labels = month) +
    ylab("number of clinical cases") +
    scale_fill_manual(values=termSeasonPalette) +
    geom_rect(data=classDates, inherit.aes=FALSE,
            aes(xmin=classesStart, xmax=classesEnd,
                ymin=-Inf, ymax=Inf,
                fill = season), alpha=.5) +
    geom_point() +
    geom_line(col="gray", aes(x=time.month.end, y=clinical.cases)) +
    geom_segment(aes(x=time.month.start, xend=time.month.end, y=clinical.cases)) +
    ggtitle("NCA&T clinical cases") +
    theme_minimal()
```

```{r}
ggsave("../output/NCAT_clinical_cases.pdf", height = 5, width = 7)
```





## Per dorm clinical data

```{r}
ccd = read.delim("../../input/from_ncat/campus_clinical-cases/reported-pos-tests_all_2021_byWeek.txt", 
                 check.names = F) %>%
    replace(is.na(.), 0) %>%
    rename(dorm = Residence)

ccd.long = ccd %>% 
    gather(key=date, value=clinical.cases, -dorm) %>%
    mutate(date = ymd(date, tz = ""))

```


```{r}
minDate = ymd("2021-01-04", tz="") # min(c(rna$date, classDates$classesStart, ccd.long$date)) #min(rna$date) #"2021-01-04 EST"
maxDate = ymd("2021-12-20", tz="") #max(c(rna$date, classDates$classesEnd, ccd.long$date))

ccplot = ggplot(ccd.long, aes(x=date, y=clinical.cases, col=dorm)) + #, pch=dorm
    # scale_y_log10() +
    ylab("number clinical cases") +
    scale_color_manual(values=dormColors) +
    # scale_shape_manual(values=dormPch) +
    scale_fill_manual(values=termSeasonPalette) +
    xlim(minDate, maxDate) +
    geom_rect(data=classDates, inherit.aes=FALSE,
              aes(xmin=classesStart, xmax=classesEnd,
                  ymin=-Inf, ymax=Inf,
                  fill = season), alpha=.2) +
    geom_point() +
    geom_line(alpha=.2) +
    theme_minimal() +
    ggtitle("SARS-CoV-2 Clinical Cases in 2021, by dorm")
ccplot
```

```{r}
ggsave(plot=ccplot, "../output/clinical-cases_2021_by-dorm_together.pdf", width = 6, height = 4)
```

```{r}
ccplot + facet_wrap("dorm", ncol=2, scales="free_y")
```

```{r}
ggsave("../output/clinical-cases_2021_by-dorm_faceted.pdf", height = 8, width = 7)
```








## County state Campus - pairwise

### Clinical cases

campus vs county, campus vs state, state vs county

For campus use:
    file: total_clinical_cases_2021-2023.xlsx
    column: 

For state use:   
    file: TABLE_DAILY_CASE&DEATH_METRICS.xlsx
    column:Molecular (PCR) Positive Cases by Specimen Date ---> sum by month to match "total_clinical_cases_2021-2023.xlsx"
```{r}
state.cc = read.delim("../../input/from_covid19.ncdhhs.gov/TABLE_DAILY_CASE-DEATHS_METRICS.txt") %>%
    select(Date, Molecular..PCR..Positive.Cases.by.Specimen.Date) %>%
    # note the odd format: day/month/year
    mutate(date = dmy(Date, tz="")) %>%
    mutate(monthYear = paste(month(date, label=T), year(date))) %>%
    # note that many of these values have commas, ex: "1,655"
    mutate(no.commas.pcr.Positive.Cases= gsub(x=Molecular..PCR..Positive.Cases.by.Specimen.Date, pattern=",", "")) %>%
    mutate(pcr.Positive.Cases = as.numeric(no.commas.pcr.Positive.Cases))
head(state.cc)
```

Take a sum for each month.
```{r}
state.cc.month = state.cc %>%
    select(monthYear, pcr.Positive.Cases) %>%
    group_by(monthYear) %>%
    summarise(state.pcr.pos.cases.month=sum(pcr.Positive.Cases, na.rm=T)) %>%
    mutate(time.month.start=dmy(paste0("1 ", monthYear), tz=""))
head(state.cc.month)
```
Recall NCA&T clinical cases from earlier...
```{r}
# cc = read.delim("data/from_Shilpi/clinical_cases_2021-2023.txt") %>% 
#     select(month, clinical.cases) %>% 
#     filter(!is.na(month)) %>% 
#     filter(!is.na(clinical.cases))
# cc$time.month.start = floor_date(dmy(paste0("15-", cc$month)), unit = "month") %>% as.POSIXct()
# cc$time.month.end = ceiling_date(dmy(paste0("15-", cc$month)), unit = "month") %>% as.POSIXct()
head(cc,3)
```

merge state and campus data
```{r}
x = cc %>% 
    mutate(campus.clinical.cases=clinical.cases) %>% 
    select(campus.clinical.cases, time.month.start) %>%
    mutate(time.month.start = as.character(time.month.start))
y = state.cc.month %>% 
    select(state.pcr.pos.cases.month, time.month.start) %>%
    as.data.frame() %>%
    mutate(time.month.start = as.character(time.month.start))
head(x)
head(y)
m.cc = merge(x = x,
             y = y, 
             by="time.month.start") %>%
    mutate(time.month.start = ymd(time.month.start, tz=""))
head(m.cc)
```

```{r}
# ymax.state=max(m.cc$state.pcr.pos.cases.month)
# ymax.campus=max(m.cc$campus.clinical.cases)
# 
# 
# ggplot(m.cc, aes(x=time.month.start, y=state.pcr.pos.cases.month)) +
#     geom_point() + 
#     geom_line() +
#     geom_point(aes(y=campus.clinical.cases))
```

```{r}
state.campus.cc.cor.test = cor.test(x=m.cc$campus.clinical.cases, y=m.cc$state.pcr.pos.cases.month, use="complete", method="spearman")

m.cc$year = as.character(year(m.cc$time.month.start))

stateVcampus = ggplot(m.cc, aes(x=campus.clinical.cases, 
                                y=state.pcr.pos.cases.month, 
                                color=year, pch=year)) +
    scale_color_manual(values=yearColors) +
    scale_shape_manual(values=yearPch) +
    scale_y_continuous(labels = comma) +
    geom_point() +
    xlab("Campus") +
    ylab("State") +
    ggtitle("Clinical Cases") +
    labs(caption=paste0("Spearmans rho: ", round(state.campus.cc.cor.test$estimate, 3),
                        "\nCor test p-value: ", round(state.campus.cc.cor.test$p.value, 3),
                        "\nN: ", nrow(m.cc))) +
    theme_minimal()
stateVcampus
```

A state-wide spike is reflected on campus, not all campus spikes are reflected at the state level.

```{r}
stateVcampusLog = stateVcampus +
           scale_x_log10() + 
           scale_y_log10() +
           labs(subtitle="log scale")

stateVcampusLog
```


```{r}
ggsave(plot=stateVcampus, "../output/clinical-cases_state-vs-campus.pdf", height = 4, width = 5)

ggsave(plot=stateVcampusLog, 
       "../output/clinical-cases_state-vs-campus_logScale.pdf", height = 4, width = 5)
```






