---
title: "PMMoV"
output: html_document
date: "2024-08-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Revisit slide 3.  We see tight correlation between Aggie V4 AS and Grab.
We have PMMoV values for AS and Grab.  If we were to normalize the SARS values by these PMMoV values, would that improve the cross-method agreement (ie correlation)?
If that improves the correlation, that makes an argument in favor of using PMMoV.
If that makes it worse, or no difference, then PMMoV is not needed.


# Setup

Libraries.
```{r}
R.Version()$version.string
tellme <- function(name){message("Package ", name, " version: ", packageVersion(name))}
library(lubridate); tellme("lubridate")
library(dplyr); tellme("dplyr")
library(tidyr); tellme("tidyr")
library(ggplot2); tellme("ggplot2")
library(Hmisc); tellme("Hmisc")
library(ggrepel); tellme("ggrepel")

# library(scales)
```

```{r}
# much of this is taken from
# http://www.sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization

#functions
tidyMelt <- function(cormat, values_name="values"){
    cormat.new = cormat %>% data.frame() 
    cormat.new$var1 = row.names(cormat.new)
    melted_cormat = cormat.new %>% pivot_longer(cols=-var1, names_to = "var2", values_to = values_name)
    return(melted_cormat)
}

make_heatmap_1 <- function(cor.co2){
    ggplot(data = cor.co2, aes(x=var1, y=var2, fill=cor)) + 
        geom_tile(color = "white")+
        scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                             midpoint = 0, limit = c(-1,1), space = "Lab", 
                             name="Correlation",
                             na.value = gray(.9)) +
        theme_minimal()+ # minimal theme
        xlab("") +
        ylab("") +
        theme(axis.text.x = element_text(angle = 90, vjust = .5,
                                         size = 8, hjust = 1)) +
        coord_fixed()
}
```


```{r}
outDir="../output"
dir.create(outDir)
```

Set color constants for the year.
```{r}
# based on ggplot defaults
yearColors = c('2021'="#00BA38", '2022'="#F8766D", '2023'="#619CFF")
yearPch = c('2021'=15, '2022'=16, '2023'=17)
```



# Intro

Note on methods:
AggieV4 hall and Barbee.
grab sampling -- all non-AS samples
compound sampling --> AS auto-sampling

## PMMoV by dorm



```{r}
pmmD.2022 = read.delim("../../input/from_ncat/PMMoV/PMMoV_2022_by-dorm_2024-07-08.txt")
pmmD.2023 = read.delim("../../input/from_ncat/PMMoV/PMMoV_2023_by-dorm_2024-07-08.txt")

# only 2022 has data for Aggie.V6, and only a few dates; drop the Aggie.V6 columns.
iNames = intersect(names(pmmD.2022), names(pmmD.2023))

pmmD = rbind(pmmD.2022[,iNames], pmmD.2023[,iNames]) %>%
    gather(key="dorm.rep", value="PMMoV", -date) %>%
    mutate(date = mdy(date, tz="")) %>%
    mutate(dorm=sub(x=dorm.rep, pattern=".rep[123]",replacement="")) %>%
    mutate(replicate=gsub(x=dorm.rep, pattern="[^?]*rep", replacement="rep")) %>%
    select(date, PMMoV, dorm, replicate) %>%
    spread(key=replicate, value=PMMoV) 

head(pmmD, 3)
```

```{r}
pmmD$mean = apply(pmmD[,c("rep1", "rep2", "rep3")], MARGIN = 1, FUN=mean)
pmmD$median = apply(pmmD[,c("rep1", "rep2", "rep3")], MARGIN = 1, FUN=median)
pmmD$sd = apply(pmmD[,c("rep1", "rep2", "rep3")], MARGIN = 1, FUN=sd)
head(pmmD, 3)
```

A little sanity check... does rep1 predict rep2 (and other pairings?).  It looks like once in a while one rep is very different.  Insead of averaging this in, we could take the median value across the three replicates.  This probably gives us a pretty similar value...right?

```{r}
ggplot(pmmD, aes(x=rep1, y=rep2)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()
ggplot(pmmD, aes(x=rep3, y=rep2)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()
ggplot(pmmD, aes(x=rep1, y=rep3)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()

ggplot(pmmD, aes(x=mean, y=median)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()
```

As expected, replicates have very similar values and taking either the median or the mean should give us very similar results.


Compare this to the per-dorm RNA copies for SARS-CoV-2.

```{r}
pmmD = pmmD %>% 
    mutate(PMMoV = median) %>%
    select(date, dorm, PMMoV) %>%
    spread(key=dorm, value=PMMoV)

methodDorms = c("Aggie.V4", "Aggie.V4.AS", "Barbee", "Barbee.AS")
pmmD_methods = pmmD %>% select(Aggie.V4, Aggie.V4.AS, Barbee, Barbee.AS)
```



## SARS by dorm



```{r}
rna.wide = read.delim("../../input/from_ncat/SARS-CoV-2_RNA-copies/RNA_copies_2021-2023.txt") %>%
    mutate(date = mdy(date, tz="")) 
rna = rna.wide %>% 
    gather(key=dorm, value=RNA.copies, Aggie.V4:Aggie.V6) %>%
    filter(!is.na(RNA.copies))
rna.wide = spread(rna, key=dorm, value=RNA.copies) #new values are numeric
head(rna)
```

```{r}
rna.wide.stringy = rna.wide %>% 
    mutate(day = day(date)) %>%
    mutate(month = month(date, label=T)) %>%
    mutate(year = as.character(year(date))) %>%
    mutate(dateString=paste(day, month, year, sep="-"))
```

```{r}
# this trick stolen from: 'Tom Constant's answer here: https://stackoverflow.com/questions/10762287/how-can-i-format-axis-labels-with-exponents-with-ggplot2-and-scales
scientific_10 <- function(x) {
    ret = x
    if (any(x > 10 | x < 0.01, na.rm = T)){
        ret = parse(text=gsub("e[+]*", " %*% 10^", scales::scientific_format()(x)))
    }
    return(ret)
}

plotASvsGrab <- function(data, AScolumn, GrabColumn, name, subtitle="RNA copies per liter",
                         imageFileName=paste0("../output/sampleMethod_", name, ".png")){
    fdata = data.frame(ASvalues=data[,AScolumn],
                       GrabValues=data[,GrabColumn],
                       dateString=data[,"dateString"],
                       year=data[,"year"])
    
    corTest = cor.test(x=fdata$GrabValues, y=fdata$ASvalues, use="complete", method="spearman")
    testN = sum(!is.na(fdata$GrabValues) & !is.na(fdata$ASvalues))
    
    ggplot(fdata, 
           aes(x=GrabValues, y=ASvalues, label=dateString, col=year, pch=year)) +
        scale_y_log10(label=scientific_10) +
        scale_x_log10(label=scientific_10) +
        scale_color_manual(values=yearColors) +
        scale_shape_manual(values=yearPch) +
        xlab("Grab Sampling") +
        ylab("Auto Sampling") +
        ggtitle(paste(name, "- comparative sampleing methods"), sub=subtitle) +
        geom_abline(slope=1, intercept = 0, alpha=.2) +
        geom_point() +
        geom_text_repel(show.legend = F) +
        labs(caption = paste0("Spearmans rho: ", round(corTest$estimate, 3), 
                              "\n", "p-value: ", corTest$p.value, "; \u2212log10(p): ", -log10(corTest$p.value),
                              "\n", "Gray line is unity (x=y).", 
                              "\n", "N=", testN)) +
        theme_classic()
    ggsave(imageFileName, width = 7, height = 7)
    return(list(corTest=corTest,
                testN=testN,
                plot=last_plot()))
}
```

```{r}
results.aggieV4.raw = plotASvsGrab(data=rna.wide.stringy, AScolumn="Aggie.V4", GrabColumn="Aggie.V4.AS", name="Aggie V4")
results.barbee.raw = plotASvsGrab(data=rna.wide.stringy, AScolumn="Barbee", GrabColumn="Barbee.AS", name="Barbee")

results.aggieV4.raw$plot
results.barbee.raw$plot
```


# normed by pmm

```{r}
methodDorms = c("Aggie.V4", "Aggie.V4.AS", "Barbee", "Barbee.AS")

rna.wide.stringy.normed = rna.wide.stringy

for (dormM in methodDorms){
    pmmDormM = pmmD[, c("date", dormM)]
    names(pmmDormM) = c("date", "PMMoV")
    
    miniStringy = rna.wide.stringy[, c("date", dormM)]
    m = merge(pmmDormM, miniStringy)
    m$normed = log10(m[,dormM]) / log10(m[,"PMMoV"])
    m2 = m[,c("date", "normed")]
    names(m2)[2] = paste0(dormM, "_normed")
    
    rna.wide.stringy.normed = merge(rna.wide.stringy.normed, m2, by="date", all = T)
}
```

```{r}
results.aggieV4.normed = plotASvsGrab(data=rna.wide.stringy.normed, AScolumn="Aggie.V4_normed", GrabColumn="Aggie.V4.AS_normed", 
                                      name="Aggie V4 by PMMoV", subtitle = "RNA copoies per liter / PMMoV copies per liter")
results.barbee.normed = plotASvsGrab(data=rna.wide.stringy.normed, AScolumn="Barbee_normed", GrabColumn="Barbee.AS_normed", 
                                     name="Barbee by PMMoV", subtitle = "RNA copoies per liter / PMMoV copies per liter")

results.aggieV4.normed$plot
results.barbee.normed$plot
```

# conclusion

For Aggie V4, The sample size went from `r results.aggieV4.raw$testN` to `r results.aggieV4.normed$testN`; and the rho when from `r results.aggieV4.raw$corTest$estimate` to `r results.aggieV4.normed$corTest$estimate`

For Barbee, The sample size went from `r results.barbee.raw$testN` to `r results.barbee.normed$testN`; and the rho when from `r results.barbee.raw$corTest$estimate` to `r results.barbee.normed$corTest$estimate`


# done

```{r}
sessionInfo()
```


