---
title: "NCA&T waste water data review"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Setup

Libraries.
```{r}
R.Version()$version.string
tellme <- function(name){message("Package ", name, " version: ", packageVersion(name))}
library(lubridate); tellme("lubridate")
library(dplyr); tellme("dplyr")
library(tidyr); tellme("tidyr")
library(ggplot2); tellme("ggplot2")
library(Hmisc); tellme("Hmisc")
library(ggrepel); tellme("ggrepel")

library(scales)
```

```{r}
# much of this is taken from
# http://www.sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization

#functions
tidyMelt <- function(cormat, values_name="values"){
    cormat.new = cormat %>% data.frame() 
    cormat.new$var1 = row.names(cormat.new)
    melted_cormat = cormat.new %>% pivot_longer(cols=-var1, names_to = "var2", values_to = values_name)
    return(melted_cormat)
}

make_heatmap_1 <- function(cor.co2){
    ggplot(data = cor.co2, aes(x=var1, y=var2, fill=cor)) + 
        geom_tile(color = "white")+
        scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                             midpoint = 0, limit = c(-1,1), space = "Lab", 
                             name="Correlation",
                             na.value = gray(.9)) +
        theme_minimal()+ # minimal theme
        xlab("") +
        ylab("") +
        theme(axis.text.x = element_text(angle = 90, vjust = .5,
                                         size = 8, hjust = 1)) +
        coord_fixed()
}
```


```{r}
# this one is from chatGPT
# Function to calculate number of pairwise complete observations for each pair of variables
calculate_pairwise_complete <- function(data) {
  num_vars <- ncol(data)
  complete_counts <- matrix(0, nrow = num_vars, ncol = num_vars)
  
  for (i in 1:num_vars) {
    for (j in i:num_vars) {
      complete_counts[i, j] <- sum(complete.cases(data[, c(i, j)]))
      if (i != j) {
        complete_counts[j, i] <- complete_counts[i, j]  # Ensure symmetry
      }
    }
  }
  
  rownames(complete_counts) <- colnames(complete_counts) <- names(data)
  return(complete_counts)
}
```

```{r}
# put a correlation matrix in long form without duplicate values
gatherCorMatrix <- function(corMat){
    corUpperTri = corMat
    corUpperTri[!upper.tri(corUpperTri)] = NA
    corLong = corUpperTri %>%
        as.data.frame() %>%
        mutate(rowName=row.names(corUpperTri)) %>%
        gather(key=colName, value=cor, -rowName) %>%
        filter(!is.na(cor))
    return(corLong)
}
```

### color constants

Some data is given by dorm.  Set some color constants for the dorm buildings.
```{r}
dormColors = c(Aggie.V4="#6A3D9A",
               Aggie.V4.AS="#CAB2D6",
               Aggie.V1="gold4",
               Aggie.V3="gray20",
               Aggie.VE="#B15928",
               Aggie.V6="#FF7F00",
               Barbee="#33A02C",
               Barbee.AS="#B2DF8A",
               Morrow="#FB9A99",
               Holland="#FDBF6F",
               Haley="#E31A1C",
               Cooper="#1F78B4",
               Pride="#A6CEE3")
dormPch = 1:length(dormColors)
names(dormPch) = names(dormColors)
```

Set color constants for spring and fall terms.
```{r}
termSeasonPalette=c(fall="gold", spring="darkolivegreen3")
```

Set color constants for the year.
```{r}
# based on ggplot defaults
yearColors = c('2021'="#00BA38", '2022'="#F8766D", '2023'="#619CFF")
yearPch = c('2021'=15, '2022'=16, '2023'=17)
```


Set other color constants.
```{r}
sarsColor = "coral"
pmmColor = "dodgerblue"
ratioColor = "pink"
rnaTypeColorPalette = c(sarsColor = "coral", pmmColor="dodgerblue", ratioColor="pink")
```

Output
```{r}
outDir="../output"
if (!dir.exists(outDir)) dir.create(outDir)
```


# NCA&T Reference

I got the dates for the first and last day of classes for spring and fall sessions for the terms corresponding to the NCA&T data.

```{r}
classDates = read.delim("../../input/from_www/ncat_class-sessions.txt", comment.char = "#") %>%
    select(term, season, classesStart, classesEnd) %>%
    # when parsing the date, use tz="" so the resulting date object is a POSIXct object to be compatible with ggplot2
    mutate(classesStart = dmy(classesStart, tz=""))  %>%
    mutate(classesEnd = dmy(classesEnd, tz=""))
classDates
```


## NCAT clinical cases

```{r}
cc = read.delim("../../input/from_ncat/campus_clinical-cases/clinical_cases_2021-2023.txt") %>%
    select(month, clinical.cases) %>%
    filter(!is.na(month)) %>%
    filter(!is.na(clinical.cases))
cc$time.month.start = floor_date(dmy(paste0("15-", cc$month), tz=""), unit = "month") %>% as.POSIXct()
cc$time.month.end = ceiling_date(dmy(paste0("15-", cc$month), tz=""), unit = "month") %>% as.POSIXct()
head(cc,3)
```

## PMMoV data

```{r}
pmm = read.delim("../../input/from_ncat/PMMoV/PMMoV_SARS_2022-2023.txt") %>%
    select(date, SARS.CoV.2.means, PMMoV.means, ratio.SARStoPMMoV) %>%
    mutate(date = mdy(date, tz="")) %>%
    mutate(logRatioS2P = log10(SARS.CoV.2.means) / log10(PMMoV.means))
head(pmm, 3)
```

### PMMoV vs SARS scatter

```{r}
pmm.vs.sars.cor.test = cor.test(x=pmm$PMMoV.means, y=pmm$SARS.CoV.2.means, method="spearman")

ggplot(pmm %>% mutate(year=as.character(year(date))), 
       aes(x=PMMoV.means, y=SARS.CoV.2.means, label=date, color=year)) +
    scale_color_manual(values=yearColors) +
    scale_x_log10() + 
    scale_y_log10() +
    geom_point() +
    ggtitle("Wastewater RNA Concentration", sub="RNA Copies per Liter") + 
    xlab("PMMoV") +
    ylab("SARS-CoV-2") + 
    theme_minimal() +
    labs(caption=paste0("Spearmans rho: ", round(pmm.vs.sars.cor.test$estimate, 3),
                        "\nCor test p-value: ", round(pmm.vs.sars.cor.test$p.value, 3),
                        "\nN: ", nrow(pmm)))
```

```{r}
ggsave("../output/PMMoV-vs-SAR_means.png", height = 4, width = 5)
```


### PMMoV and SARS over time

```{r}
# # A few constants
# sarsColor = "coral"
# pmmColor = "dodgerblue"
# ratioColor = "pink"

ggplot(pmm, aes(x=date, y=SARS.CoV.2.means)) +
    scale_y_log10() +
    geom_point(col=sarsColor) +
    theme_minimal()
ggplot(pmm, aes(x=date, y=PMMoV.means)) +
    scale_y_log10() +
    geom_point(col=pmmColor) +
    theme_minimal()
```

Put both data types together, and split using color or facets.
```{r}
pmmLong = pmm %>% tidyr::gather(key=Virus, value=Concentration, SARS.CoV.2.means, PMMoV.means, logRatioS2P)

jointPlot = ggplot(pmmLong, aes(x=date, y=Concentration, col=Virus)) +
    theme_minimal() +
    scale_y_log10() +
    geom_point() +
    geom_line() +
    scale_color_manual(values=c(SARS.CoV.2.means=sarsColor, PMMoV.means=pmmColor, logRatioS2P=ratioColor))

jointPlot
ggsave("../output/jointPlot.png")

jointPlot + facet_wrap("Virus", ncol=1, scales="free_y")
ggsave("../output/jointPlot2.png")
```

Question 1: is SARS raw data a better match for campus clinical data than the SARS/PMM data?




## PMMoV by dorm

I wonder if our total lack of corralation between PMMoV and SARS may actually be because we have collapsed the data by across unrelated samples.  So I asked Shilpi for the per-drom data.

```{r}
pmmD.2022 = read.delim("../../input/from_ncat/PMMoV/PMMoV_2022_by-dorm_2024-07-08.txt")
pmmD.2023 = read.delim("../../input/from_ncat/PMMoV/PMMoV_2023_by-dorm_2024-07-08.txt")

# only 2022 has data for Aggie.V6, and only a few dates; drop the Aggie.V6 columns.
iNames = intersect(names(pmmD.2022), names(pmmD.2023))

pmmD = rbind(pmmD.2022[,iNames], pmmD.2023[,iNames]) %>%
    tidyr::gather(key="dorm.rep", value="PMMoV", -date) %>%
    mutate(date = mdy(date, tz="")) %>%
    mutate(dorm=sub(x=dorm.rep, pattern=".rep[123]",replacement="")) %>%
    mutate(replicate=gsub(x=dorm.rep, pattern="[^?]*rep", replacement="rep")) %>%
    select(date, PMMoV, dorm, replicate) %>%
    tidyr::spread(key=replicate, value=PMMoV) 

head(pmmD, 3)
```

```{r}
pmmD$mean = apply(pmmD[,c("rep1", "rep2", "rep3")], MARGIN = 1, FUN=mean)
pmmD$median = apply(pmmD[,c("rep1", "rep2", "rep3")], MARGIN = 1, FUN=median)
pmmD$sd = apply(pmmD[,c("rep1", "rep2", "rep3")], MARGIN = 1, FUN=sd)
head(pmmD, 3)
```

A little sanity check... does rep1 predict rep2 (and other pairings?).  It looks like once in a while one rep is very different.  Insead of averaging this in, we could take the median value across the three replicates.  This probably gives us a pretty similar value...right?

```{r}
ggplot(pmmD, aes(x=rep1, y=rep2)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()
ggplot(pmmD, aes(x=rep3, y=rep2)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()
ggplot(pmmD, aes(x=rep1, y=rep3)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()

ggplot(pmmD, aes(x=mean, y=median)) +
    scale_y_log10() +
    scale_x_log10() +
    geom_abline(slope=1, col="gray") +
    geom_point()
```

As expected, replicates have very similar values and taking either the median or the mean should give us very similar results.


Compare this to the per-dorm RNA copies for SARS-CoV-2.

```{r}
pmmD = pmmD %>% mutate(PMMoV = median) 
```


Question... is PMMoV generally higher or lower in any one dorm?
```{r}
pmmd.wide = pmmD %>% select(date, dorm, PMMoV) %>% 
    tidyr::spread(key=dorm, value=PMMoV)
row.names(pmmd.wide) = pmmd.wide$date
pmmd.wide = select(pmmd.wide, -date)

# show they are highly and significantly correlated
pmmd.corList.spearman = rcorr(pmmd.wide %>% as.matrix(),type="spearman")

pmmd.pairedTpvals = data.frame(matrix(nrow=ncol(pmmd.wide), ncol=ncol(pmmd.wide)))
names(pmmd.pairedTpvals) = names(pmmd.wide)
row.names(pmmd.pairedTpvals) = names(pmmd.wide)
pmmd.TTpvals = pmmd.pairedTpvals
for (i in 1:(ncol(pmmd.wide)-1)){
    for (j in (i+1):ncol(pmmd.wide)){
        iname = names(pmmd.wide)[i]
        jname = names(pmmd.wide)[j]
        pmmd.pairedTpvals[iname,jname] = t.test(pmmd.wide[,iname], pmmd.wide[,jname], paired=T)$p.value
        pmmd.TTpvals[iname,jname] = t.test(pmmd.wide[,iname], pmmd.wide[,jname], paired=F)$p.value
    }
}

pmmPm = merge(pmmd.pairedTpvals %>% 
                  mutate(otherDorm = row.names(pmmd.pairedTpvals)) %>%
                  gather(key=dorm, value="paired.Ttest.pvalue", -otherDorm),
              pmmd.TTpvals %>% 
                  mutate(otherDorm = row.names(pmmd.TTpvals)) %>%
                  gather(key=dorm, value="Ttest.pvalue", -otherDorm)) %>% 
    filter(!is.na(Ttest.pvalue)) 

pmmPm.long = gather(pmmPm, key=test, value=p.value, -dorm, -otherDorm)
sum(pmmPm.long$p.value < 0.05)

hist(pmmPm$paired.Ttest.pvalue, breaks=20, main="PMMoV between dorms - paired t-test", xlab="p-value", col="lightblue")
hist(pmmPm$Ttest.pvalue, breaks=20, main="PMMoV between dorms - t-test", xlab="p-value", col="pink")

# ggplot(data=pmmPm, aes(x=p.value, col=test)) +
#     geom_freqpoly(binwidth=0.05)
# ggplot(data=pmmPm, aes(x=p.value, col=test)) +
#     geom_histogram(binwidth=0.05)
```

No. There were no instances (seeing 0 above) of any pair of dorms that had a significant difference in the PMMoV measurements using the t-test (paired or not).


## RNA copies


```{r}
rna.wide = read.delim("../../input/from_ncat/SARS-CoV-2_RNA-copies/RNA_copies_2021-2023.txt") %>%
    mutate(date = mdy(date, tz="")) 
rna = rna.wide %>% 
    gather(key=dorm, value=RNA.copies, Aggie.V4:Aggie.V6) %>%
    filter(!is.na(RNA.copies))
rna.wide = spread(rna, key=dorm, value=RNA.copies) #new values are numeric
head(rna)
```

This is a table of ``r nrow(rna.wide)`` rows, with dates ranging from ``r min(rna$date)`` to ``r max(rna$date)``.

### PMMoV vs SARS per dorm 

```{r}
pmm.merged = merge(x=rna %>% rename(SARS=RNA.copies),
                   y=pmmD %>% select(date, dorm, PMMoV))
head(pmm.merged,3)
```

```{r}
pmmSars.cor.test = cor.test(pmm.merged$PMMoV, pmm.merged$SARS, method="spearman")
pmmSars.cor.test.N = pmm.merged %>% filter(!is.na(PMMoV) & !is.na(SARS)) %>% nrow()
pmm.merged = mutate(pmm.merged, year = as.character(year(date)))

ggplot(pmm.merged, aes(x=PMMoV, y=SARS, 
                       color=year, pch=year)) +
    scale_x_log10() +
    scale_y_log10() +
    geom_point() +
    scale_color_manual(values=yearColors) +
    scale_shape_manual(values=yearPch) +
    ggtitle("All samples (AS and Grab)") + 
    theme_classic() +
    # theme(legend.position="none") +
    labs(caption = paste0("Spearmans rho: ", round(pmmSars.cor.test$estimate, 3), 
                          "\n", "p-value: ", round(pmmSars.cor.test$p.value,3),
                          "\n", "Gray line is unity (x=y).", 
                          "\n", "N=", pmmSars.cor.test.N)) 

ggsave("../output/PMMoV-vs-SARS_by-sample_withLegend.png", width = 6, height = 5)

ggplot(pmm.merged, aes(x=PMMoV, y=SARS, color=as.character(date))) +
    scale_x_log10() +
    scale_y_log10() +
    geom_point() +
    ggtitle("All samples (AS and Grab)") + 
    theme_minimal() +
    theme(legend.position="none") +
    labs(caption = paste0("Spearmans rho: ", round(pmmSars.cor.test$estimate, 3), 
                          "\n", "p-value: ", round(pmmSars.cor.test$p.value,3),
                          "\n", "Gray line is unity (x=y).", 
                          "\n", "N=", pmmSars.cor.test.N)) 

last_plot() + scale_colour_manual(values=c('2023-03-22'="blue",
                                 '2023-03-14'="green",
                                 '2023-01-05'="purple",
                                 '2023-01-31'="red"),
                        na.value = "grey90") +
    ggtitle(sub="All samples (AS and Grab)", "highlight random dates")

ggplot(pmm.merged %>% filter(dorm == "Aggie.V4.AS" | dorm == "Barbee.AS")
       , aes(x=PMMoV, y=SARS, color=dorm)) +
    scale_x_log10() +
    scale_y_log10() +
    geom_point() +
    ggtitle("only AS samples") +
    theme_minimal() 
# +
#     theme(legend.position="none")

ggplot(pmm.merged %>% filter(dorm == "Aggie.V4.AS" | dorm == "Barbee.AS")
       , aes(x=date, y=PMMoV, color=dorm)) +
    scale_y_log10() +
    geom_point() +
    ggtitle("only AS samples") +
    theme_minimal()

ggplot(pmm.merged %>% filter(dorm != "Aggie.V4.AS" & dorm != "Barbee.AS")
       , aes(x=PMMoV, y=SARS, color=as.character(date))) +
    scale_x_log10() +
    scale_y_log10() +
    geom_point() +
    ggtitle("only \"grab\" samples") +
    theme_minimal() +
    theme(legend.position="none")

ggplot(pmm.merged, aes(x=date, y=log10(SARS)/log10(PMMoV), color=as.character(date))) +
    geom_point() +
    theme_minimal() +
    ggtitle("Ratio") +
    theme(legend.position="none")

ggplot(pmm.merged, aes(x=date, y=log10(SARS), color=as.character(date))) +
    geom_point() +
    theme_minimal() +
    ggtitle("SARS-raw") +
    theme(legend.position="none")
```

```{r}

```



...is there...correlation within date?  
...um no.  histogram of per-date p-values is sparse but looks like uniform distribution.  How about within dorm? ...nope.  So... don't bother showing this unless someone asks for it.
```{r}
pmm.m.days = split(pmm.merged, f=pmm.merged$date)
cor.by.date = lapply(pmm.m.days, FUN=function(df){
    df = filter(df, !is.na(SARS) & !is.na(PMMoV))
    if (nrow(df) < 5) {
        return(NA)
    }
    cor.test(df$PMMoV, df$SARS, method="spearman")
})

pvals = sapply(cor.by.date, "[", "p.value")
hist(unlist(pvals), breaks=20, main="SARS vs PMMoV within a date")
abline(v=0.05)

pmm.m.dorms = split(pmm.merged, f=pmm.merged$dorm)
cor.by.dorm = lapply(pmm.m.dorms, FUN=function(df){
    df = filter(df, !is.na(SARS) & !is.na(PMMoV))
    if (nrow(df) < 5) {
        return(NA)
    }
    cor.test(df$PMMoV, df$SARS, method="spearman")
})

pvals = sapply(cor.by.dorm, "[", "p.value")
hist(unlist(pvals), breaks=20, main="SARS vs PMMoV within a dorm")
abline(v=0.05)
```

### Grab Sampling

For the rest, we'll ignore the AS data, so that we only represent each residence once.

```{r}
missAggieV4 = sum(is.na(rna.wide$Aggie.V4) & !is.na(rna.wide$Aggie.V4.AS))
missAggieV4
missBarbee = sum(is.na(rna.wide$Barbee) & !is.na(rna.wide$Barbee.AS))
missBarbee
```

There are ``r missAggieV4`` instances (dates) where we don't have an Aggie.V4 sample, but we could have one if we used pulled the AS for that dorm. As it is, Aggie.V4 has ``r sum(!is.na(rna.wide$Aggie.V4))`` data points.

There are ``r missBarbee`` instances (dates) where we don't have an Barbee sample, but we could have one if we used pulled the AS for that dorm.  As it is, Barbee has ``r sum(!is.na(rna.wide$Barbee))`` data points.

Make a new rna and rna.wide without the .AS samples.
```{r}
rna = rna %>% filter(dorm != "Aggie.V4.AS") %>% filter(dorm != "Barbee.AS")
rna.wide = rna.wide %>% select(-Aggie.V4.AS, -Barbee.AS)
```



```{r}
rcplot = ggplot(rna, aes(x=date, y=RNA.copies, col=dorm)) +
    scale_y_log10() +
    ylab("RNA copies/L") +
    scale_color_manual(values=dormColors) +
    scale_fill_manual(values=termSeasonPalette) +
    geom_rect(data=classDates, inherit.aes=FALSE,
            aes(xmin=classesStart, xmax=classesEnd,
                ymin=min(rna$RNA.copies)/10, ymax=max(rna$RNA.copies)*10,
                fill = season), alpha=.2) +
    geom_point() +
    geom_line(alpha=.2) +
    theme_minimal() +
    ggtitle("SARS-CoV-2 RNA copies over time")
rcplot
```

```{r}
ggsave("../output/SARS-CoV-2_RNA-copies-over-time_1.png")
```


```{r fig.height=20}
# TODO: save, don't show
rcplot + facet_wrap("dorm", ncol=1, scale="free_y")
ggsave("../output/SARS-CoV-2_RNA-copies-over-time_1_faceted.png", height = 12, width = 6)
```

### dorm and date -  linear model

```{r}
dorm.lm = lm(data=rna, formula = RNA.copies ~ date + dorm)
an1 = anova(dorm.lm)
p1=an1$`Pr(>F)`[1]
p2=an1$`Pr(>F)`[2]
an1
```

When we treat date as a continuous variable (as a date), its significance in the model is ``r p1``, and dorm has a significance of ``r p2``.  Seeing 0.06 and 0.5, respectively, those are both poor contributors to our model.

```{r}
rna2 = rna %>% mutate(date = as.factor(date))
dorm.lm2 = lm(data=rna2, formula = RNA.copies ~ date + dorm)
an2 = anova(dorm.lm2)
an2.p1=an2$`Pr(>F)`[1]
an2.p2=an2$`Pr(>F)`[2]
an2
```

When we treat date as a factor, its significance in the model is ``r an2.p1``, and dorm has a significance of ``r an2.p2``. Seeing e-16 and 0.3, we can say that date matters, but dorm does not.

(Results were near identical using glm.)


Lets do a leave one out prediction by dorm, and see how it does.
```{r}
dorms = names(dormColors) %>% grep(pattern="AS", invert = T, value = T)

rna.ml = rna %>% 
    mutate(RNA.copies.Log10 = log10(RNA.copies)) 

lou.plot.list = list()
for (dorm.i in dorms){
    trainData = rna.ml %>% filter(dorm != dorm.i) %>% select(date, RNA.copies.Log10)
    testData = rna.ml %>% filter(dorm == dorm.i)
    testIn = testData %>% select(date)
    testOut = testData %>% select(RNA.copies.Log10)
    
    # make the model
    model = lm(trainData, formula = RNA.copies.Log10 ~ date)
    
    # predict
    testData$prediction = predict(model, testIn)
    
    # plot
    lou.plot.list[[dorm.i]] <- ggplot(data=testData, aes(x=RNA.copies.Log10, y=prediction)) +
        # scale_y_log10() +
        # scale_x_log10() +
        scale_color_manual(values=dormColors) +
        geom_abline(slope=1, intercept = 0, alpha=.2) +
        geom_point() +
        ggtitle(dorm.i) +
        theme_minimal()
}
lou.plot.list[[1]]
lou.plot.list[[2]]
lou.plot.list[[3]]
lou.plot.list[[4]]
lou.plot.list[[5]]
```



### Per dorm clinical data

```{r}
ccd = read.delim("../../input/from_ncat/campus_clinical-cases/reported-pos-tests_all_2021_byWeek.txt",
                 check.names = F) %>%
    replace(is.na(.), 0) %>%
    rename(dorm = Residence)

ccd.long = ccd %>% 
    gather(key=date, value=clinical.cases, -dorm) %>%
    mutate(date = ymd(date, tz = ""))
```

```{r}
minDate = min(c(rna$date, classDates$classesStart, ccd.long$date))
maxDate = ymd("2021-12-20", tz="") #max(c(rna$date, classDates$classesEnd, ccd.long$date))

ccplot = ggplot(ccd.long, aes(x=date, y=clinical.cases, col=dorm)) +
    # scale_y_log10() +
    ylab("number clinical cases") +
    scale_color_manual(values=dormColors) +
    scale_fill_manual(values=termSeasonPalette) +
    xlim(minDate, maxDate) +
    geom_rect(data=classDates, inherit.aes=FALSE,
              aes(xmin=classesStart, xmax=classesEnd,
                  ymin=-Inf, ymax=Inf,
                  fill = season), alpha=.2) +
    geom_point() +
    geom_line(alpha=.2) +
    theme_minimal() +
    ggtitle("SARS-CoV-2 Clinical Cases in 2021, by dorm")
ccplot

```

```{r}
ggsave(plot=ccplot, "../output/clinical-cases_2021_by-dorm_together.png", width = 6, height = 4)
```

```{r}
ccplot + facet_wrap("dorm", ncol=2, scales="free_y")
```

```{r}
ggsave("../output/clinical-cases_2021_by-dorm_faceted.png", height = 8, width = 7)
```



### RNA copies and case counts per dorm together

```{r}
rna.tmp = rna %>%
    arrange(date) %>%
    mutate(week = round_date(date, unit="week")) %>%
    select(-date)

# Save a table of the ways this COULD  be summarized to 1-value-per-week
rna.tmp1 = summarise(.data=rna.tmp, .by=c(week, dorm), n=length(RNA.copies))
rna.tmp2 = summarise(.data=rna.tmp, .by=c(week, dorm), min=min(RNA.copies))
rna.tmp3 = summarise(.data=rna.tmp, .by=c(week, dorm), max=max(RNA.copies))
rna.tmp4 = summarise(.data=rna.tmp, .by=c(week, dorm), mean=mean(RNA.copies))
rna.week.summary = merge(merge(rna.tmp1, rna.tmp2),
                         merge(rna.tmp3, rna.tmp4)) %>%
    arrange(desc(n))
write.table(rna.week.summary, file="../output/summarize_RNA-copies_per-week.txt", 
            sep="\t", quote=F, row.names=F)
plot(log10(rna.week.summary$mean), log10(rna.week.summary$max))

# move forward using the max value for each week for each dorm
rna.week = rna.tmp %>% summarise(.by=c(week, dorm), RNA.copies=max(RNA.copies))


rna.cc.d = merge(rna.week, ccd.long %>% rename(week=date), by=c("dorm", "week")) %>%
    filter(complete.cases(RNA.copies, clinical.cases))
```

```{r}
ggplot(data=rna.week.summary %>% mutate(year = factor(year(week))), 
       aes(x=min, y=max, size=n, col=year)) + 
    geom_point(alpha=.4) +
    scale_x_log10() +
    scale_y_log10() +
    geom_point(data=rna.week.summary %>% filter(week=="2023-01-08"), 
               col="black", shape=1, size=5,
               aes(x=min, y=max)) +
    labs(title ="Summaries within a dorm, across measures from the same week",
         caption="Black circles highlight values from week=2023-01-08.")


```

```{r}
rna %>% 
    mutate(week = round_date(date, unit="week")) %>% 
    filter(week=="2023-01-08") %>% 
    arrange(date)
```


```{r}
peakWeeks = rna %>%
    mutate(week = round_date(date, unit="week")) %>% 
    filter(RNA.copies > 1e11) 

for (pw in unique(peakWeeks$week)){
    weeksTable = rna %>% 
               mutate(week = round_date(date, unit="week")) %>% 
               filter(week==pw) %>% 
               arrange(date) %>%
               mutate(date = as.character(date))
    ggplot(data=weeksTable,
           aes(x=date, y=RNA.copies)) +
        geom_boxplot(outlier.shape = NA) +
        geom_jitter(height = 0, aes(color=dorm)) +
        scale_y_log10() +
        ggtitle(unique(weeksTable$week))
    show(last_plot())
}

unique(peakWeeks$week)
```
Follow up on the one for week "2023-04-16". Expand to the week before.
```{r}
weeksTable = rna %>% 
    mutate(week = round_date(date, unit="week")) %>% 
    filter(week=="2023-04-16" | week=="2023-04-09") %>% 
    arrange(date) %>%
    mutate(date = as.character(date))
ggplot(data=weeksTable,
       aes(x=date, y=RNA.copies)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(height = 0, aes(color=dorm)) +
    scale_y_log10() 
```

```{r}
weeksTable = rna %>% 
    mutate(week = round_date(date, unit="week")) %>% 
    filter(week=="2022-01-02" | week=="2022-01-09" | week=="2022-01-16" | week=="2022-01-23") %>% 
    arrange(date) %>%
    mutate(date = as.character(date))
ggplot(data=weeksTable,
       aes(x=date, y=RNA.copies)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(height = 0, aes(color=dorm)) +
    scale_y_log10() 

weeksTable = rna %>% 
    mutate(week = round_date(date, unit="week")) %>% 
    filter(week=="2021-01-03" | week=="2021-01-10" | week=="2021-01-17" | week=="2021-01-24") %>% 
    arrange(date) %>%
    mutate(date = as.character(date))
ggplot(data=weeksTable,
       aes(x=date, y=RNA.copies)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(height = 0, aes(color=dorm)) +
    scale_y_log10() 
```




Scatter plots

```{r}
cc.vs.sars.cor.test = cor.test(x=rna.cc.d$RNA.copies, y=rna.cc.d$clinical.cases, method="spearman")

scatPlot = ggplot(rna.cc.d, aes(x=RNA.copies, y=clinical.cases, color=dorm)) +
    scale_x_log10() +
    xlab("RNA copies (max value in a given week)") +
    ylab("clinical cases (total over a given week)") +
    labs(title="SARS-CoV-2 2021", subtitle = "by week and dorm") +
    scale_color_manual(values=dormColors) +
    geom_point() +
    theme_minimal()

scatPlot.together = scatPlot +
    labs(caption=paste0("Spearmans rho: ", round(cc.vs.sars.cor.test$estimate, 3),
                        "\nCor test p-value: ", round(cc.vs.sars.cor.test$p.value, 3),
                        "\nN: ", nrow(rna.cc.d)))
scatPlot.together
```
```{r}
ggsave(scatPlot.together, filename="../output/scatter_clinical-cases_vs_RNA.copies_perDorm_together_max.png", width = 5, height = 5)
```

...and if we had used mean instead...
```{r}
# move forward using the max value for each week for each dorm
rna.week = rna.tmp %>% summarise(.by=c(week, dorm), RNA.copies=max(RNA.copies))


rna.cc.d = merge(rna.week, ccd.long %>% rename(week=date), by=c("dorm", "week")) %>%
    filter(complete.cases(RNA.copies, clinical.cases))

rm(cc.vs.sars.cor.test)
cc.vs.sars.cor.test = cor.test(x=rna.cc.d$RNA.copies, y=rna.cc.d$clinical.cases, method="spearman")

scatPlot = ggplot(rna.cc.d, aes(x=RNA.copies, y=clinical.cases, color=dorm)) +
    scale_x_log10() +
    xlab("RNA copies (mean value in a given week)") +
    ylab("clinical cases (total over a given week)") +
    labs(title="SARS-CoV-2 2021", subtitle = "by week and dorm") +
    scale_color_manual(values=dormColors) +
    geom_point() +
    theme_minimal()

scatPlot.together = scatPlot +
    labs(caption=paste0("Spearmans rho: ", round(cc.vs.sars.cor.test$estimate, 3),
                        "\nCor test p-value: ", round(cc.vs.sars.cor.test$p.value, 3),
                        "\nN: ", nrow(rna.cc.d)))
scatPlot.together

ggsave(scatPlot.together, filename="../output/scatter_clinical-cases_vs_RNA.copies_perDorm_together_mean.png", width = 5, height = 5)
```



Spearman test per dorm.
```{r}
rna.cc.d.split = split(rna.cc.d, f=rna.cc.d$dorm)
rna.cc.d.split = rna.cc.d.split[sapply(rna.cc.d.split, nrow) > 2]
dorm.cors = lapply(rna.cc.d.split,
                   FUN=function(df){
                       print(nrow(df))
                       return( cor.test(x=df$RNA.copies, y=df$clinical.cases, method="spearman") )
                   })
dorm.stat.table = data.frame(N = sapply(rna.cc.d.split, nrow),
                             p.value = sapply(dorm.cors, "[[", "p.value"),
                             rho = sapply(dorm.cors, "[[", "estimate"))
write.table(cbind(dorm=row.names(dorm.stat.table), dorm.stat.table), "../output/spearman_clinical-cases_vs_RNA.copies_perDorm.txt",
            sep="\t", row.names = F, quote=F)
dorm.stat.table
```



```{r}
scatPlot + facet_wrap("dorm", ncol=3)
```
```{r}
ggsave("../output/scatter_clinical-cases_vs_RNA.copies_perDorm_faceted.png", width = 7, height = 8)
```


```{r}
head(rna.cc.d)
```

```{r}
linmod = lm(data=rna.cc.d, formula= clinical.cases ~ RNA.copies)
anova(linmod)
```

```{r}
linmod = lm(data=rna.cc.d, formula= clinical.cases ~ RNA.copies * dorm)
anova(linmod)
```

```{r}
linmod = lm(data=rna.cc.d, formula= clinical.cases ~ RNA.copies / dorm)
anova(linmod)
```

```{r}
ggplot(data=rna.cc.d, aes(x=dorm, y=clinical.cases, color=dorm)) +
    geom_boxplot() +
    scale_color_manual(values=dormColors)

ggplot(data=rna.cc.d, aes(x=dorm, y=RNA.copies, color=dorm)) +
    geom_boxplot() +
    scale_color_manual(values=dormColors) +
    scale_y_log10()
```


```{r}
library(nlme)
m.mixed = lme(data=rna.cc.d, clinical.cases ~ RNA.copies, random= ~1 | dorm)
anova(m.mixed)
```


# Data from covid19.ncdhhs.gov



## Percent Positive cases


<!-- ## County state Campus - pairwise -->

<!-- ### Clinical cases -->

<!-- campus vs county, campus vs state, state vs county -->

<!-- For campus use: -->
<!--     file: total_clinical_cases_2021-2023.xlsx -->
<!--     column:  -->

<!-- For state use:    -->
<!--     file: TABLE_DAILY_CASE&DEATH_METRICS.xlsx -->
<!--     column:Molecular (PCR) Positive Cases by Specimen Date ---> sum by month to match "total_clinical_cases_2021-2023.xlsx" -->
<!-- ```{r} -->
<!-- state.cc = read.delim("data/from_covid19.ncdhhs.gov/TABLE_DAILY_CASE-DEATHS_METRICS.txt") %>% -->
<!--     select(Date, Molecular..PCR..Positive.Cases.by.Specimen.Date) %>% -->
<!--     # note the odd format: day/month/year -->
<!--     mutate(date = dmy(Date, tz="")) %>% -->
<!--     mutate(monthYear = paste(month(date, label=T), year(date))) %>% -->
<!--     # note that many of these values have commas, ex: "1,655" -->
<!--     mutate(no.commas.pcr.Positive.Cases= gsub(x=Molecular..PCR..Positive.Cases.by.Specimen.Date, pattern=",", "")) %>% -->
<!--     mutate(pcr.Positive.Cases = as.numeric(no.commas.pcr.Positive.Cases)) -->
<!-- head(state.cc) -->
<!-- ``` -->

<!-- Take a sum for each month. -->
<!-- ```{r} -->
<!-- state.cc.month = state.cc %>% -->
<!--     select(monthYear, pcr.Positive.Cases) %>% -->
<!--     group_by(monthYear) %>% -->
<!--     summarise(state.pcr.pos.cases.month=sum(pcr.Positive.Cases, na.rm=T)) %>% -->
<!--     mutate(time.month.start=dmy(paste0("1 ", monthYear), tz="")) -->
<!-- head(state.cc.month) -->
<!-- ``` -->
<!-- Recall NCA&T clinical cases from earlier... -->
<!-- ```{r} -->
<!-- # cc = read.delim("data/from_Shilpi/clinical_cases_2021-2023.txt") %>%  -->
<!-- #     select(month, clinical.cases) %>%  -->
<!-- #     filter(!is.na(month)) %>%  -->
<!-- #     filter(!is.na(clinical.cases)) -->
<!-- # cc$time.month.start = floor_date(dmy(paste0("15-", cc$month)), unit = "month") %>% as.POSIXct() -->
<!-- # cc$time.month.end = ceiling_date(dmy(paste0("15-", cc$month)), unit = "month") %>% as.POSIXct() -->
<!-- head(cc,3) -->
<!-- ``` -->

<!-- merge state and campus data -->
<!-- ```{r} -->
<!-- x = cc %>%  -->
<!--     mutate(campus.clinical.cases=clinical.cases) %>%  -->
<!--     select(campus.clinical.cases, time.month.start) %>% -->
<!--     mutate(time.month.start = as.character(time.month.start)) -->
<!-- y = state.cc.month %>%  -->
<!--     select(state.pcr.pos.cases.month, time.month.start) %>% -->
<!--     as.data.frame() %>% -->
<!--     mutate(time.month.start = as.character(time.month.start)) -->
<!-- head(x) -->
<!-- head(y) -->
<!-- m.cc = merge(x = x, -->
<!--              y = y,  -->
<!--              by="time.month.start") %>% -->
<!--     mutate(time.month.start = ymd(time.month.start, tz="")) -->
<!-- head(m.cc) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # ymax.state=max(m.cc$state.pcr.pos.cases.month) -->
<!-- # ymax.campus=max(m.cc$campus.clinical.cases) -->
<!-- #  -->
<!-- #  -->
<!-- # ggplot(m.cc, aes(x=time.month.start, y=state.pcr.pos.cases.month)) + -->
<!-- #     geom_point() +  -->
<!-- #     geom_line() + -->
<!-- #     geom_point(aes(y=campus.clinical.cases)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- state.campus.cc.cor.test = cor.test(x=m.cc$campus.clinical.cases, y=m.cc$state.pcr.pos.cases.month, use="complete", method="spearman") -->

<!-- m.cc$year = as.character(year(m.cc$time.month.start)) -->

<!-- stateVcampus = ggplot(m.cc, aes(x=campus.clinical.cases,  -->
<!--                                 y=state.pcr.pos.cases.month,  -->
<!--                                 color=year)) + -->
<!--     scale_color_manual(values=yearColors) + -->
<!--     scale_y_continuous(labels = comma) + -->
<!--     geom_point() + -->
<!--     xlab("Campus") + -->
<!--     ylab("State") + -->
<!--     ggtitle("Clinical Cases") + -->
<!--     labs(caption=paste0("Spearmans rho: ", round(state.campus.cc.cor.test$estimate, 3), -->
<!--                         "\nCor test p-value: ", round(state.campus.cc.cor.test$p.value, 3), -->
<!--                         "\nN: ", nrow(m.cc))) + -->
<!--     theme_minimal() -->
<!-- stateVcampus -->
<!-- ``` -->

<!-- A state-wide spike is reflected on campus, not all campus spikes are reflected at the state level. -->

<!-- ```{r} -->
<!-- stateVcampusLog = stateVcampus + -->
<!--            scale_x_log10() +  -->
<!--            scale_y_log10() + -->
<!--            labs(subtitle="log scale") -->

<!-- stateVcampusLog -->
<!-- ``` -->


<!-- ```{r} -->
<!-- ggsave(plot=stateVcampus, "../output/clinical-cases_state-vs-campus.png", height = 4, width = 5) -->

<!-- ggsave(plot=stateVcampusLog,  -->
<!--        "../output/clinical-cases_state-vs-campus_logScale.png", height = 4, width = 5) -->
<!-- ``` -->


### Campus case counts vs RNA.copies --- section moving to sep markdown

Get RNA copies per month across the 3-year period
```{r}
rna.wide.month = rna.wide %>% mutate(month=round_date(date, unit="month"))

rna.long.month = rna.wide.month %>% 
    select(-date) %>% 
    gather(key=dorm, value=RNA.copies, -month) %>%
    filter(!is.na(RNA.copies)) %>%
    group_by(month) %>%
    summarise(mean.RNA.copies=mean(RNA.copies), 
              meanLog.RNA.copies=mean(log10(RNA.copies)), 
              max.RNA.copies=max(RNA.copies), 
              median.RNA.copies=median(RNA.copies), 
              min.RNA.copies=min(RNA.copies), 
              n.RNA.copies.vals=n()) %>%
    mutate(year=as.factor(lubridate::year(month)))

ggplot(rna.long.month, aes(x=mean.RNA.copies, y=max.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")
ggplot(rna.long.month, aes(x=mean.RNA.copies, y=median.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")
ggplot(rna.long.month, aes(x=min.RNA.copies, y=median.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")

ggplot(rna.long.month, aes(x=mean.RNA.copies, y=meanLog.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")
ggplot(rna.long.month, aes(x=max.RNA.copies, y=meanLog.RNA.copies, size=n.RNA.copies.vals, color=year)) +
    geom_point(alpha=.5) +
    scale_x_log10() +
    scale_y_log10() +
    ggtitle("Summary metrics across dorms, across weeks within a month")

```

Merge in campus case counts by month, across the three year period.
```{r}
cc.month = cc %>% select(-month) %>% rename(month = time.month.start)
campus.cc.rna = merge(cc.month, 
                      rna.long.month, by="month")
head(campus.cc.rna)
```

```{r}
rm(ct)
ct=cor.test(x=campus.cc.rna$mean.RNA.copies, y=campus.cc.rna$clinical.cases, use="complete", method="spearman")
ggplot(data=campus.cc.rna, aes(x=mean.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()

rm(ct)
ct=cor.test(x=campus.cc.rna$median.RNA.copies, y=campus.cc.rna$clinical.cases, use="complete", method="spearman")
ggplot(data=campus.cc.rna, aes(x=median.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()

rm(ct)
ct=cor.test(x=campus.cc.rna$max.RNA.copies, y=campus.cc.rna$clinical.cases, use="complete", method="spearman")
ggplot(data=campus.cc.rna, aes(x=max.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()
ggsave("../output/campus-wide-cc_vs_max-rna-copies.pdf")

rm(ct)
ct=cor.test(x=campus.cc.rna$meanLog.RNA.copies, y=campus.cc.rna$clinical.cases, use="complete", method="spearman")
ggplot(data=campus.cc.rna, aes(x=meanLog.RNA.copies, y=clinical.cases, color=year)) +
    geom_point() +
    scale_x_log10() +
    ggtitle("NCA&T campus wide values") +
    labs(caption=paste0("Spearmans rho: ", round(ct$estimate, 3),
                        "\nCor test p-value: ", round(ct$p.value, 3),
                        "\nN: ", nrow(campus.cc.rna))) +
    theme_minimal()
```

```{r}

```


```{r}
sessionInfo()
```






